The state learner \\ {-- a super learner for right-censored data}

### 2024-05-21 tex/draft-version/state-learner-submit-version0.tex
renaming

\documentclass[a4,danish]{article}
\usepackage[a4paper, margin=3.5cm]{geometry}
\input{./latex-settings/standard-settings.tex}
\input{./latex-settings/standard-commands.tex}
\input{./latex-settings/theorem-env.tex}
\input{./latex-settings/title-compact.tex}

@@

%% Specific commands
\newcommand{\data}{\ensuremath{\mathcal{D}}}

\begin{document}

% \articletype{ARTICLE TEMPLATE}% Specify the article type or omit as appropriate

@@

\author{Anders Munch and Thomas A. Gerds
  % \name{Anders Munch\textsuperscript{a}\thanks{CONTACT: Anders Munch. Email: a.munch@sund.ku.dk}
  %   and Thomas A. Gerds\textsuperscript{a}}
  % \affil{\textsuperscript{a} Section of Biostatistics, Department of Public Health, University of Copenhagen}
}

\maketitle

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. Unfortunately, the
  commonly used partial likelihood loss is not suited for super learning, and
  inverse probability of censoring weighted loss functions require a
  pre-specified estimator of the censoring distribution. To relax this, we
  introduce the state learner, a new super learner for survival analysis, which
  evaluates the loss based on the observed data simultaneously using libraries
  of predictions models for the event(s) of interest and the censoring
  distribution. We establish an oracle inequality for the state learner and
  investigate its performance through numerical experiments. We illustrate how
  the state learner allows us to estimate causal effects in a competing risks
  setting without having to pre-specify models for neither the cause-specific
  hazards of interest nor the censoring distribution.
  % In survival analysis, prediction models are needed as stand-alone tools and in
  % applications of causal inference to estimate nuisance parameters. The super
  % learner is a machine learning algorithm which combines a library of prediction
  % models into a meta learner based on cross-validated loss. Unfortunately, the
  % commonly used partial likelihood loss is not suited for super learning, and
  % inverse probability of censoring weighted loss functions require a
  % pre-specified estimator of the censoring distribution. To relax this, we
  % introduce the state learner, a new super learner for survival analysis, which
  % evaluates the loss based on the observed data simultaneously for two libraries
  % of prediction models, one for the event time distribution and one for the
  % censoring distribution. We establish an oracle inequality for the state
  % learner and investigate its performance through numerical experiments.
\end{abstract}

@@

The article is organized as follows. We introduce our notation and framework in
Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
learning in general with right-censored data, and in
Section~\ref{sec:super-learner-simple} we introduce the state learner.
Section~\ref{sec:theor-results-prop} provides theoretical guarantees for the
state learner. In Section~\ref{sec:targeted-learning} we discuss the use of the
state learner in the context of targeted learning. We report a numerical study
in Section~\ref{sec:numer-exper}, and analyze a prostate cancer data set in
Section~\ref{sec:real-data-appl}. Finally, we relate the state learner to
existing approaches in Section~\ref{sec:discussion} and discuss some limitations
of our proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.

@@

@@

The problem with most existing super learners for right-censored data is that
they depend on a pre-specified estimator of the censoring distribution. The main
idea of the state learner is to jointly use learners of \( \Lambda_1 \),
\( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
distribution \( P \). The discrete state learner ranks a tuple of learners of
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
observed data. To formally introduce the state learner, we define the
multi-state process
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde D=2\}+ 3\,\1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau].
\end{equation*}
Note that at time \(t\), we observe that each observation is in one of four
mutually exclusive states (Figure \ref{fig:multi-state-process}).
\begin{figure}[h]
  \centering
  \includegraphics[width=.5\textwidth]{../figures/figure-multi-state-process.pdf}
  \caption{Illustration of the multi-state process \(\eta\) used by
    the state learner. Note that `censored' is a state, hence the
    process is always observed at any time.}
  \label{fig:multi-state-process}
\end{figure}
The conditional distribution of \( \eta(t) \) given \( X=x \) is
determined by the function
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
  \quad \text{for all} \quad
  t \in [0,\tau],
  k \in \{0,1,2,3\},
  x \in \mathcal{X}.
\end{equation}
The function \( F \) describes the conditional state occupation
probabilities corresponding to the observed multi-state process
\(\eta\).

@@

An example of a targeted estimator is the \fxnote*{also TMLE, depending on
  implementation}{one-step estimator}, defined as
\begin{equation}
    \label{eq:one-step-def}
    \hat{\Psi}_{\text{OS}}(\data_n)
    =
    \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
    H(\data_n))
    + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
      \hat{b}_n(\data_n) )]},
\end{equation}
where \( \empmeas \) is the empirical measure of a sample \(\{O_i\}_{i=1}^n\).
We can make the following asymptotic expansion of the one-step estimator
\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric,rytgaard2022continuous},
\begin{equation*}
  \hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)
  =  \empmeas{[\psi(\blank ; \Lambda_1, \Lambda_2, \Gamma)]}
  +\mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
\end{equation*}
where the remainder term is typically such that
\begin{equation}
  \label{eq:4}
  \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
  = \mathcal{O}_P{
    \left\{
      (\Lambda_1-\hat{\Lambda}_{1,n})^2
      +
      (\Lambda_2-\hat{\Lambda}_{2,n})^2
      +
      (\Gamma-\hat{\Gamma}_{n})^2
    \right\}
  }.
\end{equation}
Hence, under a suitable Donsker class regularity condition
\citep{bickel1993efficient,kennedy2016semiparametric}, when
equation~(\ref{eq:4}) holds and the nuisance parameters $\Lambda_1$,
$\Lambda_2$, and $\Gamma$ are consistently estimated at rate
\( \smallO_P{(n^{-1/4})} \), then
\begin{equation}
  \label{eq:3}
  \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}).
\end{equation}
In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
obtain asymptotically valid \((1-\alpha)\cdot100\%\) confidence intervals by
calculating
\begin{equation*}
  \left[
    \hat{\Psi}_{\text{OS}}(\data_n) - q_{\alpha/2} \hat{\sigma}(\data_n) ,
    \;
    \hat{\Psi}_{\text{OS}}(\data_n) + q_{\alpha/2} \hat{\sigma}(\data_n)
  \right],
\end{equation*}
where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile or the standard normal
distribution, and
\begin{equation*}
  \hat{\sigma}(\data_n) = 
  \left(
    \empmeas{
      \left[
        \psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n))^2
      \right]}
  \right)^{1/2}.
\end{equation*}

@@

\subsection{Cause-specific average treatment effect}
\label{sec:cause-spec-aver}

We now demonstrate how the state learner and the general estimation strategy
outlined above can be used to estimate a concrete parameter of interest in the
competing risk setting. We assume that the covariate vector \( X \in \R^p \) can
be separated into a binary treatment indicator \( A \) and a vector of potential
confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We abuse notation slightly
by writing \( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
\( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \). We use $\mu$ to denote
the marginal distribution of \( W \) and $\pi$ to denote the conditional
probability of treatment,
\begin{equation*}
  \pi(w) = P(A=1 \mid W=w).
\end{equation*}
We assume throughout that $\pi$ is uniformly bounded away from \( 0 \) and
\( 1 \) on \( \mathcal{W} \). As both \( A \) and \( W \) are fully observed for
all individuals we can use a super learner to estimate $\pi$
\fxnote*{add}{[ref]}, and we denote this estimator by $\hat{\pi}_n$. We use the
empirical measure of \( \{W_1, \dots, W_n\} \) to estimate $\mu$, and denote
this estimator by the $\hat{\mu}_n$. As parameter of interest we consider the
\fxnote*{what is this called?}{standardized difference in absolute risk of cause
  \( 1 \) before time $\tau$},
\begin{equation*}
  \theta_{{\tau}}(Q) = \int_{\mathcal{W}} 
  \left\{
    \int_0^{\tau}
    S(s- \mid w, 1)  \Lambda_1(\diff s \mid w, 1)
    -
    \int_0^{\tau}
    S(s- \mid w, 0)  \Lambda_1(\diff s \mid w, 0)
  \right\}
  \mu(\diff w).
\end{equation*}
Under a set of
additional assumptions, \( \theta_{{\tau}} \) can be given the causal
interpretation
\begin{equation*}
  \theta_{{\tau}}(Q) =
  \E{[T^{A=1} \leq \tau, D^{A=1}=1]}-
  \E{[T^{A=0} \leq \tau, D^{A=0}=1]},
\end{equation*}
where \( (T^A, D^A) \) denotes potential outcomes \citep{hernanRobinsWhatIf}. In
this case, the interpretation of $\theta_{\tau}$ is the difference in the
average risk of cause \( 1 \) occurring before time \( \tau \) in the population
if everyone had been given treatment (\( A=1 \)) compared to if no one had been
given treatment.

@@

Equations~(\ref{eq:1}) and~(\ref{eq:5}) allow us to construct a one-step
estimator by using the definition given in equation~(\ref{eq:one-step-def}),
which gives the estimator
\begin{equation}
  \label{eq:one-step-comp-ate}
  \begin{split}
    \hat{\Psi}_{t,\text{OS}}(\data_n)
    = &
        \tilde{\Psi}_{t}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
        \hat{\mu}_n(\data_n))
    \\
      &
        +
        \empmeas{[\psi_{\tau}(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
    \\
    = &
        \empmeas{[\psi_{\tau}^1(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
    \\
      &
        - \empmeas{[\psi_{\tau}^0(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
  \end{split}
\end{equation}

@@

%% IPCW with potentially misspecified censoring model.
%%
%% Applying SL, concrete, and Westling on classical data set. (Where to put
%% this?)

We compare the state learner with two other discrete super learners and an
oracle selector in a simulation study without a competing event. The two other
super learners are based on inverse probability of censoring weighted (IPCW)
Brier scores \citep{graf1999assessment,gerds2006consistent}, and we refer to
these as IPCW super learners. These super learners depend on an estimator of the
censoring distribution, and we consider IPCW super learners that use either the
Kaplan-Meier estimator (IPCW(KM)) or a correctly specified Cox model (IPCW(Cox))
to estimate the censoring distribution. Both IPCW super learners are fitted
using the \texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}. Each discrete super learner provides a
learner for the cumulative hazard function for the outcome of interest, and from
this a risk prediction model can be obtained. We measure performance of each
super learner in terms of the Brier score of the provided risk prediction model
at a specific time horizon. The Brier score is approximated using a large
(\( n = 20,000 \)) independent data set of uncensored data. The oracle selector
uses the large data set of uncensored event times to select the learner with the
lowest expected Brier score. The expected Brier score of the oracle selector
serves as a lower benchmark value. For all super learners we split the data into
five folds for training and testing.

Note that given a learner for the cumulative hazard function of the
outcome event, we can typically use the same method to construct a
learner of the cumulative hazard function of the censoring
distribution. This would typically work by training the learner on the
data set \( \data_n^c \), where \( \data_n^c = \{O_i^c\}_{i=1}^n \)
with \( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we
use a learner for the cumulative hazard function of the outcome to
learn the cumulative hazard function of the censoring time, we mean
that the learner is trained on \( \data_n^c \).

\paragraph{Scenario 1}

We first generate a simple dataset to demonstrate that an IPCW super learner can
perform poorly when the censoring model is misspecified. We start by generating a
binomial baseline covariate \( A \) with success probability \( 30\% \). We then
generate outcome and censoring variables according to a Cox-Weibull distribution
with hazard rates of approximately \( 0.5 \) and \( 2.5 \), respectively. We use
a library \( \mathcal{A} \) consisting of two learners, the (marginal)
Kaplan-Meier estimator and the Kaplan-Meier estimator stratified on the binary
covariate. For the state learner we use the same learners to construct the
library \( \mathcal{B} \).

The results are shown in Figure~\ref{fig:ipcw-fail}. We see that the IPCW super
learner based on a misspecified censoring model (IPCW(KM)) has a larger Brier
score than the other super learners and that its performance does not improve
much with sample size. % When the censoring model is misspecified, the
% IPCW(KM) super learner fails to identify the learner corresponding to the
% data-generating mechanism, even though this learner is included in the
% library.
The performance of the state learner is comparable to or slightly better than
the IPCW super learner based on a correctly pre-specified censoring model. The
performance of both the IPCW(Cox) super learner and the state learner are close
to that of the oracle for most sample sizes.

\begin{figure}
  \centerline{\includegraphics[width=1\linewidth]{../figures/ipcw-fail.pdf}}
  \caption[]{Results for scenario~1 of the simulation study. For the learner
    selected by each of the four discrete super learners, the Brier score
    calculated in a large independent data set without censoring is plotted
    against sample size. The results are based on 200 repetitions.}
  \label{fig:ipcw-fail}
\end{figure}

\paragraph{Scenario 2}

We next generate data according to a more complex distribution motivated from a
real dataset in which censoring depends on the baseline covariates. We simulate
data based on a prostate cancer study described in
\citep{kattan2000pretreatment}. The outcome of interest was the time to tumor
recurrence, and five baseline covariates were used to predict outcome:
prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values between 6
and 10), radiation dose (RD), hormone therapy (HT, yes/no) and clinical stage
(CS, six values). The study was designed such that a patient's radiation dose
depended on when the patient entered the study \citep{gerds2013estimating}. This
in turn implied that the time of censoring depended on the radiation dose.
% It was found that almost all covariates were predictive of censoring.
The data were re-analyzed in \citep{gerds2013estimating} where a sensitivity
analysis was conducted based on simulated data. We use the same simulation
setup, where event and censoring times are generated according to parametric
Cox-Weibull models estimated from the original data, and the covariates are
generated according to either marginal Gaussian normal or binomial distributions
estimated from the original data
\citep[c.f.,][Section~4.6]{gerds2013estimating}. We use the library consisting
of the nine learners described in Table~\ref{tab:zel-library}. For the state
learner we use the same library to learn the censoring distribution.

>>>>

\documentclass[a4,danish]{article}
\usepackage[a4paper, margin=3.5cm]{geometry}
\input{./latex-settings/standard-settings.tex}
\input{./latex-settings/standard-commands.tex}
\input{./latex-settings/theorem-env.tex}
\input{./latex-settings/title-compact.tex}
\definecolor{linkcolor}{rgb}{0, 0, 0.54}

@@

\newcommand{\data}{\ensuremath{\mathcal{D}}}

@@

\author{Anders Munch and Thomas A. Gerds}

\begin{document}
\maketitle

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. Unfortunately, the
  commonly used partial likelihood loss is not suited for super learning, and
  inverse probability of censoring weighted loss functions require a
  pre-specified estimator of the censoring distribution. To relax this, we
  introduce the state learner, a new super learner for survival analysis, which
  evaluates the loss based on the observed data simultaneously using libraries
  of predictions models for the event(s) of interest and the censoring
  distribution. We establish an oracle inequality for the state learner and
  investigate its performance through numerical experiments. We illustrate how
  the state learner allows us to estimate causal effects in a competing risks
  setting without having to pre-specify models for neither the cause-specific
  hazards of interest nor the censoring distribution.
\end{abstract}

@@

The article is organized as follows. We introduce our notation and framework in
Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
learning in general with right-censored data, and in
Section~\ref{sec:relat-liter-exist} we review existing methods.
Section~\ref{sec:super-learner-simple} introduces our novel proposal, the state
learner, and Section~\ref{sec:theor-results-prop} provides theoretical
guarantees for the state learner. In Section~\ref{sec:targeted-learning} we
discuss the use of the state learner in the context of targeted learning. We
report some numerical experiments in Section~\ref{sec:numer-exper} and analyze a
prostate cancer data set in Section~\ref{sec:real-data-appl}.
Section~\ref{sec:discussion} contains a discussion of the merits and limitations
of our proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.

@@

\section{Existing methods}
\label{sec:relat-liter-exist}

Machine learning based on right-censored data commonly use the negative partial
log-likelihood as loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
However, this loss function is unsuited for super learning, because many
canonical survival learners (e.g., the Kaplan-Meier estimator, random survival
forest, and semi-parametric Cox models) provide cumulative hazard functions that
are piece-wise constant in the time argument, and hence assign zero probability
to event times not observed in the training data. This implies that when data
are observed in continuous time, any of these learners will almost surely have
infinite loss in any independent hold-out sample according to the negative
partial log-likelihood loss.

When a proportional hazards model is assumed, the baseline hazard function can
be profiled out of the likelihood to give a new partial log-likelihood loss
\citep{cox1972regression}, which has been suggested as a loss function for super
learning \citep{golmakani2020super,verweij1993cross}. While this allows the
library of learners to include Cox' proportional hazard models, the drawback is
that the library is now restricted to include only these models.

Another approach for super learning with right-censored data is to use an
inverse probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
An IPCW loss function is attractive because the associated risk does not depend
on the censoring distribution but describes a feature of the population of
interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
be obtained using censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
methods rely on an estimator of the censoring distribution, and their drawback
is that this estimator has to be pre-specified. When the data-generating
mechanism is complex and not well-understood, pre-specification of the censoring
distribution is a challenge.

As far as we are aware, the only existing attempt at avoiding the need to
pre-specify a censoring model is a recent proposal suggested independently by
\cite{han2021inverse} and \cite{westling2021inference}. The authors do not
consider competing risks but suggest to iterate between learning \( \Lambda \)
and $\Gamma$ using IPCW loss functions and select the final learner when the
iterative procedure has converged.
% Using the estimator \( \hat{b}^{(0)}(\data_n^c) \) of $\Gamma$ based on some
% initial learner \( \hat{b}^{(0)} \in \mathcal{B} \), we calculate the
% estimated IPCW risk for each leaner in \( \mathcal{A} \). Picking
% \( a^{(1)} \) as the discrete super learner according to this IPCW loss
% function, we then estimate weight
To the best of our knowledege, no general theoretical guarantees exist for this
procedure. It seems unclear whether this iterative procedure will always be able
to find a correctly specified model in a given library.

@@

The problem with most existing super learners for right-censored data is that
they depend on a pre-specified estimator of the censoring distribution. The main
idea of the state learner is to jointly use learners of \( \Lambda_1 \),
\( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
distribution \( P \). The discrete state learner ranks a tuple of learners of
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
observed data. To formally introduce the state learner, we define the
multi-state process
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde D=2\}+ 3\,\1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau].
\end{equation*}
Note that at time \(t\), we observe that each observation is in one of four
mutually exclusive states (Figure \ref{fig:multi-state-process}).
\begin{figure}[h]
  \centering
  \includegraphics[width=.5\textwidth]{../figures/figure-multi-state-process.pdf}
  \caption{Illustration of the multi-state process \(\eta\) used by
    the state learner. Note that `censored' is a state, hence the
    process is always observed at any time.}
  \label{fig:multi-state-process}
\end{figure}
The conditional distribution of \( \eta(t) \) given \( X=x \) is determined by
the \fxnote*{reconsider letting censoring state be \( -1 \)?}{function}
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
  \quad \text{for all} \quad
  t \in [0,\tau],
  k \in \{0,1,2,3\},
  x \in \mathcal{X}.
\end{equation}
The function \( F \) describes the conditional state occupation
probabilities corresponding to the observed multi-state process
\(\eta\).

@@

An example of a targeted estimator is the one-step estimator, defined as
\begin{equation}
    \label{eq:one-step-def}
    \hat{\Psi}_{\text{OS}}(\data_n)
    =
    \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
    H(\data_n))
    + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
      \hat{b}_n(\data_n) )]},
\end{equation}
where \( \empmeas \) is the empirical measure of a sample \(\{O_i\}_{i=1}^n\).
We can make the following asymptotic expansion of the one-step estimator
\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric,rytgaard2022continuous},
\begin{equation*}
  \hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)
  =  \empmeas{[\psi(\blank ; \Lambda_1, \Lambda_2, \Gamma)]}
  +\mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
\end{equation*}
where the remainder term is typically such that
\begin{equation}
  \label{eq:4}
  \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
  = \mathcal{O}_P{
    \left\{
      \|\Lambda_1-\hat{\Lambda}_{1,n}\|^2
      +
      \|\Lambda_2-\hat{\Lambda}_{2,n}\|^2
      +
      \|\Gamma-\hat{\Gamma}_{n}\|^2
    \right\}
  },
\end{equation}
for some suitable norm \( \|\blank \| \), for instance the
\( \mathcal{L}_{P}^2 \)-norm. Hence, under a suitable Donsker class regularity
condition \citep{bickel1993efficient,kennedy2016semiparametric}, when
equation~(\ref{eq:4}) holds and the nuisance parameters $\Lambda_1$,
$\Lambda_2$, and $\Gamma$ are consistently estimated at rate
\( \smallO_P{(n^{-1/4})} \), then
\begin{equation}
  \label{eq:3}
  \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}),
\end{equation}
where we use \( \weakly \) to denote weak convergence \citep{van2000asymptotic}.
In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
obtain asymptotically valid \((1-\alpha)\cdot100\%\) confidence intervals by
calculating
\begin{equation*}
  \left[
    \hat{\Psi}_{\text{OS}}(\data_n) - q_{\alpha/2} \hat{\sigma}(\data_n) ,
    \;
    \hat{\Psi}_{\text{OS}}(\data_n) + q_{\alpha/2} \hat{\sigma}(\data_n)
  \right],
\end{equation*}
where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile or the standard normal
distribution, and
\begin{equation*}
  \hat{\sigma}(\data_n) = 
  \left(
    \empmeas{
      \left[
        \psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n))^2
      \right]}
  \right)^{1/2}.
\end{equation*}

@@

\subsection{Average treatment effect on the cumulative incidence}
\label{sec:cause-spec-aver}

In the following we demonstrate how the state learner and the general estimation
strategy outlined above can be used to construct an estimator of the
cause-specific average treatment effect. We assume that the covariate vector
\( X \in \R^p \) can be separated into a binary treatment indicator \( A \) and
a vector of potential confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We
abuse notation slightly by writing
\( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
\( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \). We use $\mu$ to denote
the marginal distribution of \( W \) and $\pi$ to denote the conditional
probability of treatment,
\begin{equation*}
  \pi(w) = P(A=1 \mid W=w).
\end{equation*}
We assume throughout that $\pi$ is uniformly bounded away from \( 0 \) and
\( 1 \) on \( \mathcal{W} \). As both \( A \) and \( W \) are fully observed for
all individuals we can use an existing super learner to estimate $\pi$
\citep{Polley_Ledell_Kennedy_Laan_2023_Superlearn}, and we denote this estimator
by $\hat{\pi}_n$. We use the empirical measure of \( \{W_1, \dots, W_n\} \) to
estimate $\mu$, and denote this estimator by the $\hat{\mu}_n$. As parameter of
interest we consider the standardized difference in the cumulative incidence due
to cause 1 at time $\tau$, which can be expressed as
\begin{equation*}
  \theta_{{\tau}}(Q) = \int_{\mathcal{W}} 
  \left\{
    \int_0^{\tau}
    S(s- \mid w, 1)  \Lambda_1(\diff s \mid w, 1)
    -
    \int_0^{\tau}
    S(s- \mid w, 0)  \Lambda_1(\diff s \mid w, 0)
  \right\}
  \mu(\diff w).
\end{equation*}
Under a set of
additional assumptions, \( \theta_{{\tau}} \) can be given the causal
interpretation
\begin{equation*}
  \theta_{{\tau}}(Q) =
  P{(T^{A=1} \leq \tau, D^{A=1}=1)}-
  P{(T^{A=0} \leq \tau, D^{A=0}=1)},
\end{equation*}
where \( (T^A, D^A) \) denotes potential outcomes \citep{hernanRobinsWhatIf}. In
this case, the interpretation of $\theta_{\tau}$ is the difference in the
average risk of cause \( 1 \) occurring before time \( \tau \) in the population
if everyone had been given treatment (\( A=1 \)) compared to if no one had been
given treatment.

@@

Equations~(\ref{eq:1}) and~(\ref{eq:5}) allow us to construct a one-step
estimator by using the definition given in equation~(\ref{eq:one-step-def}),
which gives the estimator
\begin{equation}
  \label{eq:one-step-comp-ate}
  \begin{split}
    \hat{\Psi}_{t,\text{OS}}(\data_n)
    = &
        \tilde{\Psi}_{t}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
        \hat{\mu}_n(\data_n))
    \\
      &
        +
        \empmeas{[\psi_{\tau}(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
    \\
    = &
        \empmeas{[\psi_{\tau}^1(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
    \\
      &
        - \empmeas{[\psi_{\tau}^0(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}.
  \end{split}
\end{equation}

@@

As a proof of concept we conduct two small simulation studies in a setting
without competing risks. We generate data according to a distribution motivated
from a real dataset in which censoring depends on the baseline covariates. We
simulate data based on a prostate cancer study described in
\citep{kattan2000pretreatment}. The outcome of interest was the time to tumor
recurrence, and five baseline covariates were used to predict outcome:
prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values between 6
and 10), radiation dose (RD), hormone therapy (HT, yes/no) and clinical stage
(CS, six values). The study was designed such that a patient's radiation dose
depended on when the patient entered the study \citep{gerds2013estimating}. This
in turn implied that the time of censoring depended on the radiation dose. The
data were re-analyzed in \citep{gerds2013estimating} where a sensitivity
analysis was conducted based on simulated data. We use the same simulation
setup, where event and censoring times are generated according to parametric
Cox-Weibull models estimated from the original data, and the covariates are
generated according to either marginal Gaussian normal or binomial distributions
estimated from the original data
\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
setting as ``dependent censoring''. We also considered a simulation setting
where data were generated in the same way, except that censoring was generated
completely independently. We refer to this simulation setting as ``independent
censoring''.

For all super learners we use a library consisting of three learners: The
Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
model with main effects \citep{cox1972regression, survival-package}, and a
random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
library to learn the outcome and the censoring mechanism. To be precise, note
that given a learner for the cumulative hazard function of the outcome event, we
can typically use the same method to construct a learner of the cumulative
hazard function of the censoring distribution. This would typically work by
training the learner on the data set \( \data_n^c \), where
\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
for the cumulative hazard function of the outcome to learn the cumulative hazard
function of the censoring time, we mean that the learner is trained on
\( \data_n^c \).

We first compare the state learner to two IPCW based super learners: The first
uses a Cox model with main effects to estimate the censoring probabilities
(IPCW(Cox)), while the second uses the Kaplan-Meier estimator to estimate the
censoring probabilities (IPCW(KM)). The Cox model for the censoring distribution
is correctly specified in both simulation settings while the Kaplan Meier
estimator only estimates the censoring model correctly in the simulation setting
where censoring is independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023}.
Each super learner provides a learner for the cumulative hazard function for the
outcome of interest, and from this a risk prediction model can be obtained. We
measure performance of each super learner by calculating the index of prediction
accuracy (IPA) \citep{kattan2018index} at a fixed time horizon (36 months) for
the risk prediction model provided by the super learner. The IPA is 1 minus the
ratio between the model's Brier score and the null model's Brier score, where
the null model is the model that does not use any covariate information. The IPA
is approximated using a large (\( n = 20,000 \)) independent data set of
uncensored data. As a benchmark we calculate the performance of the risk
prediction model chosen by the oracle selector. The oracle selector uses the
large data set of uncensored event times to select the learner with the highest
IPA.

The results are shown in Figure~\ref{fig:ipcw-fail}. We see that when censoring
is not independent, using the Kaplan-Meier estimator to estimate the censoring
probabilities provides a risk prediction model with an IPA that is lower than
the risk prediction model provided by the state learner. The performance of the
risk prediction model selected by the state learner is similar to the risk
prediction model selected by the IPCW(Cox) super learner which a priori uses a
correctly specified model for the censoring distribution. Both these risk
prediction models are close to the performance of the oracle, except for small
sample sizes.

\begin{figure}
  \centerline{\includegraphics[width=1\linewidth]{../figures/experiment-fig-sl-ipcw.pdf}}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are based on
    1000 repetitions and the error bars are used to quantify the Monte Carlo
    uncertainty.}
  \label{fig:ipcw-fail}
\end{figure}

We next compare the state learner to the super learner \texttt{survSL}
\citep{westling2021inference}, as this is the only existing super learner we
know which works without a pre-specified censoring model. As both the state
learner and \texttt{survSL} also provides a prediction model for the probability
of being censored, we compare the performance of both the outcome and the
censoring modeling using the IPA.

The results are shown in Figure~\ref{fig:zelefski}. We see that the prediction
models selected for both censoring and outcome have similar or higher API
compared to the prediction models selected by \texttt{survSL} for most sample
sizes.

\begin{figure}
  \centerline{\includegraphics[width=1\linewidth]{../figures/experiment-fig-sl-survSL.pdf}}
  \caption[]{For the risk prediction models of both the outcome and the
    censoring probability provided by each of the super learners, the IPA is
    plotted against sample size. The results are based on 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski}
\end{figure}

\section{Example with prostate cancer study}
\label{sec:real-data-appl}
